Rapid long-range disynaptic inhibition explains the formation of cortical orientation maps.

### 2017-02-23 paper.tex

Competition between populations of neurons has been proposed as one of the canonical computations of cortical 
networks, and has been hypothesized to underly a range of brain functions including working memory \citep{Amit1995,Durstewitz2000},
orientation tuning \citep{Somers1995,Ben-Yishai1995}, and functional map development \citep{VonderMalsburg1973,CMVC,Antolik2011}.
In developmental models of functional cortical organization the competition occurs between populations of neurons spatially offset 
along the cortical surface, whereby local populations mutually reinforce each other via excitatory connections (short-range 
excitation) while long-range inhibition facilitates competition between the local populations and stabilizes the activity in the 
network. Such so called Mexican-hat arrangement of recurrent interactions, however, is in stark contrast to the know anatomical arrangement
of cortical circuitry: the excitatory neurons (especially in superficial layers) tend to form long-range arborizations spanning 
multiple columns while the axons of a majority of inhibitory neurons are confined to an area only several hundreds of micrometers 
in diameter \citep{Buzas2006,Budd2001}.

@@

As we will show in this study, the nature of the transmission delays between different neural type (excitatory,inhibitory) combinations
holds the key to resolving this long standing open question. A recent study by Ohana \etal\,\citep{Ohana2012} revealed a specific pattern of transmission delays between different neural type combinations. Specifically they found that the excitatory to excitatory connections are slow while the excitatory to inhibitory connections are fast. In this study we explore the 
possibility that this specific transmission delay pattern is the missing link that can explain how short range inhibition can lead 
to effective cortical competition. We employ computational models to show that the fast excitatory-to-inhibitory-to-excitatory pathway allows for the disynaptic inhibition to generate the effective Mexican-hat like lateral interactions, thus for the first time 
explaining competition in topologically organized cortical networks with no biologically implausible assumptions. To demonstrate the
proposed properties, in this study we focus on models of development of orientation maps, because this is the most well explored
example of competition driven functional cortical organization development, but these results generalize to development of other
cortical topological properties, as well as potentially other competition based computations.

>>>>

Competition between populations of neurons has been proposed as one of the canonical computations of cortical 
networks, and has been hypothesized to underly a range of brain functions including working memory \citep{Amit1995,Durstewitz2000},
orientation tuning \citep{Somers1995,Ben-Yishai1995}, and functional map development \citep{VonderMalsburg1973,CMVC,Antolik2011}.
In developmental models of functional cortical organization the competition occurs between populations of neurons spatially offset 
along the cortical surface, whereby local populations mutually reinforce each other via excitatory connections (short-range 
excitation) while long-range inhibition facilitates competition between the local populations and stabilizes the activity in the 
network. Such so called Mexican-hat arrangement of recurrent interactions (figure \ref{fig:lat_int}A), however, is in stark contrast to the know anatomical arrangement
of cortical circuitry: the excitatory neurons (especially in superficial layers) tend to form long-range arborizations spanning 
multiple columns while the axons of a majority of inhibitory neurons are confined to an area only several hundreds of micrometers 
in diameter \citep{Buzas2006,Budd2001} (figure \ref{fig:lat_int}B).

@@

\begin{figure}[htpb!] 
\centering
\includegraphics[width=16cm]{./SVG/FigureLateralInteractions/drawing.png}
\caption{Cortical anatomy and effective lateral interactions. (A) Short-range excitation (red) and long-range inhibition (blue) leads to effective Mexican-hat lateral interactions (black curve). 
(B) Anatomical evidence indicates the oposite organization of lateral connectivity in cortex, whereby excitatory neurons send long-range connections to other neurons, while inhibitory neurons only 
local ones. Furthermore, uniform delays accross the intra-cortical projections is tipically assumed. It is not clear how such anatomical configuration can support the Mexican-hat effective interaction
across cortical surface. (C) Under the assumption of slow excitatory-to-excitatory and fast excitatory-to-inhibitory and inhibitory-to-excitatory pathway (and disregarding the inhibitory-to-inhibitory interactions), 
the effective lateral inhibitory interactions correspond to the convolution of the excitatory-to-inhibitory and inhibitory-to-excitatory connection kernels. Uner the assumption that these kernels are Gaussian (G($\sigma$) in 
the figure), the effective inhibitory interactions will be Gaussian with variance (space-constant) greater then the excitatory-to-excitatory one, thus forming Mexican-hat lateral interaction profile.
}
\label{fig:lat_int}
\end{figure}

As we will show in this study, the nature of the transmission delays between different neural type (excitatory,inhibitory) combinations
holds the key to resolving this long standing open question. A recent study by Ohana \etal\,\citep{Ohana2012} revealed a specific pattern of transmission delays between different neural type combinations. Specifically they found that the excitatory to excitatory connections are slow while the excitatory to inhibitory connections are fast. In this study we explore the 
possibility that this specific transmission delay pattern is the missing link that can explain how short range inhibition can lead 
to effective cortical competition. We employ computational models to show that the fast excitatory-to-inhibitory-to-excitatory pathway allows for the disynaptic inhibition to generate the effective Mexican-hat like lateral interactions (figure \ref{fig:lat_int}C), thus for the first time 
explaining competition in topologically organized cortical networks with no biologically implausible assumptions. To demonstrate the
proposed properties, in this study we focus on models of development of orientation maps, because this is the most well explored
example of competition driven functional cortical organization development, but these results generalize to development of other
cortical topological properties, as well as potentially other competition based computations.
